---
title: "Gov 50 Final Project"
author: "Ty Schanhofer"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to voting in this year's Wisconsin Senate election. I want to look at it this year especially because of the concurrent Presidential election. I wonder if there are some interesting effects of the bigger, national election happening at the same time.
## Proposal Paragraph
https://www.kaggle.com/datasets/thedevastator/u-s-public-school-system-finances-esec14
https://www.kaggle.com/datasets/davidbroberts/us-gdp-by-state-19972020
  (I am moving away from my thoughts above) 
  
  Does funding for an American state's elementary, middle, and high schools have correlation/a causal relationship with a state's GDP per capita? I hypothesize that the greater amount of funding per student that is put towards education, the greater a state's GDP per capita. I expect states that put more money towards educating its youth will then produce a workforce that is made up of more high earners. Additionally, I believe these states will also attract high earners from elsewhere, which would also raise the GDP per capita. My explanatory value is the amount of money each state spends per student. My outcome value is the GDP per capita for each state. If I find that greater funding for schools per student is accompanied by greater GDP per capita for a state, then my hypothesis will be proven. If I find that there is little or reverse correlation between a state's funding for schools, then my hypothesis will be disproven. When I run a regression of a state's GDP with school funding as the indicator, a positive, significant coefficient would indicate support for my hypothesis.
  

```{r}
library(tidyverse)
library(ggplot2)
```

  
##Data Upload
```{r}
rawgdp <- read_csv("data/bea-gdp-by-state.csv")
rawschools <- read_csv("data/ELSI_csv_export_finance2.csv")
```

##Data Cleaning

```{r}
gdp <- rawgdp |>
  select(Fips, Area, "2017")

gdp
```



##Graphs

```{r}
ggplot(data = gdp,
       mapping = aes(x = `2017`)) +
geom_histogram() +
  labs(title = "GDP per Capita Across the US")
```



```{r}
schools <- rawschools |>
  select("Area", "Total Revenues (TR) per Pupil (MEMBR) [State Finance] 2016-17")

schools
```


```{r}
gdp
```





  