---
title: "Gov 50 Final Project"
author: "Ty Schanhofer"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to voting in this year's Wisconsin Senate election. I want to look at it this year especially because of the concurrent Presidential election. I wonder if there are some interesting effects of the bigger, national election happening at the same time.

## Proposal Paragraph
https://www.kaggle.com/datasets/thedevastator/u-s-public-school-system-finances-esec14
https://www.kaggle.com/datasets/davidbroberts/us-gdp-by-state-19972020
  (I am moving away from my thoughts above) 
  
  Does funding for an American state's elementary, middle, and high schools have correlation/a causal relationship with a state's total GDP? I hypothesize that the greater amount of gross funding that is put towards education, the greater a state's GDP. I expect states that put more money towards educating its youth will then produce a workforce that is made up of more high earners. My explanatory value is the amount of money each state's total school revenue. My outcome value is the total GDP for each state. If I find that greater funding for schools is accompanied by greater GDP for a state, then my hypothesis will be proven. If I find that there is little or reverse correlation between a state's funding for schools, then my hypothesis will be disproven. When I run a regression of a state's GDP with school funding as the indicator, a positive, significant coefficient would indicate support for my hypothesis.
  

```{r}
library(tidyverse)
library(ggplot2)
```

  
Step 1. Load in Data
```{r}
revenue <- read_csv("data/u-s-public-school-system-finances-esec14")
```

  
  
  
  
  